Online Social Cohesion Reflects Real-World Group Action in Syria during the Arab Spring

### 2020-06-02 manuscript.Rmd
Slight phrasing edit
More formatting and typos

Following a wave of democratization efforts across the Middle East and North
Africa, modest Syrian protests began in March 2011. The regime responded harshly
with imprisonments, detentions, censorship, and military operations as the
resistance to Assad’s regime grew and spread throughout the country. The
fighting escalated dramatically in March 2012, after increased unrest in the
western city of Homs. International actors---including the Red Cross/Red
Crescent Society, the United Nations General Assembly, the Security Council, and
individual world leaders---worked toward cease-fires, negotiations, and
concessions without much success. While many high-level military and diplomatic
officials defected from the regime, numerous loyal forces continued the regime's
repression, and the Syrian people turned to social media to help coordinate
their activism and communicate to the world. Between March and June 2012, more
than 8 million tweets were sent within Syria, revealing the details of the
conflict suppressed by the government.

@@

Others have similarly expressed hesitation about the role of online
organizing. Some have argued that real-life strong ties outweigh online weak ties and
that the power of the strong ties through in-person contact is critical
to recruiting people for higher-risk activities
[@gladwell2010revolution; @mcadam1986recruitment].

@@

Citizens communicate with each other with increased transparency due to the
ubiquitous availability of social media such as low-cost and egalitarian
platforms like Twitter. Citizens are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and
military incursions, the degree of state repression, and the number of
casualties. In turn, the government---interested in remaining in
power---monitors social media platforms to ascertain emerging threats
[@comninos2011twitter], such as where new groups are forming and where existing
groups are strengthening, where gatherings are held, and who the idea
entrepreneurs are. An example of a leader expressing his interest in such public
conversation is given by Bashar al-Assad in a speech on June 3, 2012:

@@

With few exceptions, Assad chose to keep social media platforms up and running.
Gohdes [-@gohdes2015pulling] notes that regimes can attempt to thwart the growth
of social protest online and in the streets by shutting down the Internet.
During these times, however, she finds that military repression increases.
However, government information about the citizens decreases; as such, many
regimes have elected to maintain Internet connectivity as a means of monitoring
and surveilling their citizens.

@@

Previous research on social movements supports the idea that participants are
likely to join movements when they know and have a personal connection to
someone else already involved [@mcadam1986recruitment]. The type of collective
contentious behavior that manifests in society is a function of the available
political opportunity structures. The political opportunity structure are
“comprised *(sic)* of specific configuration of resources, institutional
arrangements and historical precedents for social mobilization, which facilitate
the development of protest movements in some instances and constrain them in
others” [@kitschelt1986political]. What emerges as strategies and tactics in
contentious politics are, in essence, a function of what the system makes
possible. This includes the number of people and their disposition (i.e.
nonviolent or violent), as well as their internal structure. The more cohesive a
group is, the greater threat they pose to the establishment.

@@

The civil war literature has further explored the role of social cohesion in
detail. Social homogeneity increases the likelihood of social conflict,
especially when the society is polarized [@collier2004greed]; this is partly
due to nationalist groups' strong identification with a collective identity
[@cederman2007beyond] and their social solidarity [@sambanis2009s]. These
same processes can be similarly important for rebels [@anyanwu2002economic].
Referencing Gurr’s [-@gurr2000peoples] work on minorities at
risk, Regan and Norton [-@regan2005greed] note that

@@

Thus, lexical and interpersonal cohesion are hypothesized to be tied to social
and political cohesion. Shared language in an online environment may lead to the
development or expression of a shared identity [@wiltermuth2009synchrony]. This,
in turn, may drive individuals’ propensity to engage in similar coordinated
behavior on behalf of the group [@wiltermuth2009synchrony; @walton2012mere].
Taken together, these two processes may grow movement participants' weak and
strong ties [@granovetter1977strength; @gladwell2010revolution;
@mcadam1986recruitment], strengthening the movement overall by expanding and
deepening heterogenous ties. Therefore, in the context of social conflict, the
linguistic cohesion of the online conversation should correspond to the degree
of social cohesion on the ground.

Steinert-Threlkeld [-@steinert2017spontaneous] modeled a similar relationship
between social media and social mobilization alongside event data. This
groundbreaking work demonstrates that participants on the fringe of a social
movement contribute substantially to the overall level of protest. In addition
to providing new methods to this growing field of study, we build on this
previous work by demonstrating how sociolinguistic cohesion evolves in
response to negative and positive exogenous events.

@@

## Data Preparation
### Social Cohesion Metric
The social cohesion or linguistic similarity of daily tweets was calculated by
quantifying the frequency of shared content words with respect to those words’
frequencies in the entire tweet corpus. If a linguistic term has a high
frequency in one tweet but has a low frequency in the whole tweet corpus, then
it is more likely a content word and is thus given more weight in calculating
social cohesion. This method is similar to inverse document frequency (IDF). For
more information on IDF, see @jones1972statistical and
@robertson2004understanding.

@@

The social cohesion calculation was performed on the tweet corpus containing
both English and Arabic tweets. All hashtags and URLs were included in the
analysis.

The raw social cohesion time series can be seen in Figure \@ref(fig:raw-cohesion-ts).

@@

The raw event time series for target filtered data can be seen in Figure \@ref(fig:raw-event-ts-target). The The raw event time series for source and target filtered data can be seen in Figure \@ref(fig:raw-event-ts-source-target).

@@

#### Metrics
For WCRQA analyses, we decided to focus on RR and DET, but WCRQA visualization can be done for any metric desired.

>>>>

Following a wave of democratization efforts across the Middle East and North
Africa, modest Syrian protests began in March 2011. The regime responded harshly
with imprisonments, detentions, censorship, and military operations as the
resistance to Assad's regime grew and spread throughout the country. The
fighting escalated dramatically in March 2012, after increased unrest in the
western city of Homs. International actors---including the Red Cross/Red
Crescent Society, the United Nations General Assembly, the Security Council, and
individual world leaders---worked toward cease-fires, negotiations, and
concessions without much success. While many high-level military and diplomatic
officials defected from the regime, numerous loyal forces continued the regime's
repression, and the Syrian people turned to social media to help coordinate
their activism and communicate to the world. Between March and June 2012, more
than 8 million tweets were sent within Syria, revealing the details of the
conflict suppressed by the government.

@@

\noindent Others have similarly expressed hesitation about the role of online
organizing. Some have argued that real-life strong ties outweigh online weak ties and
that the power of the strong ties through in-person contact is critical
to recruiting people for higher-risk activities
[@gladwell2010revolution; @mcadam1986recruitment].

@@

Citizens communicate with each other with increased transparency due to the
ubiquitous availability of social media such as low-cost and egalitarian
platforms like Twitter. Citizens are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and
military incursions, the degree of state repression, and the number of
casualties. In turn, the government---interested in remaining in
power---monitors social media platforms to ascertain emerging threats
[@comninos2011twitter], such as where new groups are forming and where existing
groups are strengthening, where gatherings are held, and who the idea
entrepreneurs are.

An example of a leader expressing his interest in such public
conversation is given by Bashar al-Assad in a speech on June 3, 2012:

@@

\noindent With few exceptions, Assad chose to keep social media platforms up and running.
Gohdes [-@gohdes2015pulling] notes that regimes can attempt to thwart the growth
of social protest online and in the streets by shutting down the Internet.
During these times, however, she finds that military repression increases.
However, government information about the citizens decreases; as such, many
regimes have elected to maintain Internet connectivity as a means of monitoring
and surveilling their citizens.

@@

Previous research on social movements supports the idea that participants are
likely to join movements when they know and have a personal connection to
someone else already involved [@mcadam1986recruitment]. The type of collective
contentious behavior that manifests in society is a function of the available
political opportunity structures. The political opportunity structure are
“comprised of *(sic)* specific configuration of resources, institutional
arrangements and historical precedents for social mobilization, which facilitate
the development of protest movements in some instances and constrain them in
others” [@kitschelt1986political]. What emerges as strategies and tactics in
contentious politics are, in essence, a function of what the system makes
possible. This includes the number of people and their disposition (i.e.
nonviolent or violent), as well as their internal structure. The more cohesive a
group is, the greater threat they pose to the establishment.

@@

The civil war literature has further explored the role of social cohesion in
detail. Social homogeneity increases the likelihood of social conflict,
especially when the society is polarized [@collier2004greed]; this is partly
due to nationalist groups' strong identification with a collective identity
[@cederman2007beyond] and their social solidarity [@sambanis2009s]. These
same processes can be similarly important for rebels [@anyanwu2002economic].
Referencing Gurr's [-@gurr2000peoples] work on minorities at
risk, Regan and Norton [-@regan2005greed] note that

@@

Thus, lexical and interpersonal cohesion are hypothesized to be tied to social
and political cohesion. Shared language in an online environment may lead to the
development or expression of a shared identity [@wiltermuth2009synchrony]. This,
in turn, may drive individuals' propensity to engage in similar coordinated
behavior on behalf of the group [@wiltermuth2009synchrony; @walton2012mere].
Taken together, these two processes may grow movement participants' weak and
strong ties [@granovetter1977strength; @gladwell2010revolution;
@mcadam1986recruitment], strengthening the movement overall by expanding and
deepening heterogenous ties. Therefore, in the context of social conflict, the
linguistic cohesion of the online conversation should correspond to the degree
of social cohesion on the ground.

Steinert-Threlkeld [-@steinert2017spontaneous] modeled a similar relationship
between social media and social mobilization alongside event data. That
groundbreaking work demonstrated that participants on the fringe of a social
movement contribute substantially to the overall level of protest. In addition
to providing new methods to this growing field of study, we build on this
previous work by demonstrating how sociolinguistic cohesion evolves in
response to negative and positive exogenous events.

@@

## Data Preparation
### Social Cohesion Metric
The social cohesion or linguistic similarity of daily tweets was calculated by
quantifying the frequency of shared content words with respect to those words'
frequencies in the entire tweet corpus. If a linguistic term has a high
frequency in one tweet but has a low frequency in the whole tweet corpus, then
it is more likely a content word and is thus given more weight in calculating
social cohesion. This method is similar to inverse document frequency (IDF). For
more information on IDF, see @jones1972statistical and
@robertson2004understanding.

@@

\noindent The social cohesion calculation was performed on the tweet corpus
containing both English and Arabic tweets. All hashtags and URLs were included
in the analysis. The raw social cohesion time series can be seen in Figure
\@ref(fig:raw-cohesion-ts).

@@

The raw event time series for target filtered data can be seen in Figure
\@ref(fig:raw-event-ts-target). The raw event time series for source and target
filtered data can be seen in Figure \@ref(fig:raw-event-ts-source-target).

@@

#### Metrics
For WCRQA analyses, we decided to focus on RR and DET, but similar analyses and
visualizations can be performed for any RQA metric.

### 2020-06-02 manuscript.tex
More formatting and typos

\abstract{
Social mobilization requires effective collective coordination. In recent
years, individuals involved in social movements have taken to social media
platforms to reach broader audiences at rapid speeds. Effective virtual
mobilization similarly requires social cohesion and coordination, yet the
degree to which virtual mobilization reflects or drives real-world movements
is unclear. Here, we explore the dynamics of real-world events and social
cohesion on Twitter in Syria during the Arab Spring. Using the nonlinear
methods cross-recurrence quantification analysis and windowed cross-recurrence
quantification analysis, we investigate if counts of all daily events, daily
positive events, and daily negative events are coupled with social cohesion
found in Syrian tweets. Results indicate that virtual social cohesion is
coupled with the counts of all and negative events each day but shows a more
nuanced connection to positive events, only visible when examining finer-grained
data. This suggests that virtual mobilization does reflect
real-world movements, with the strongest connections to the frequency of all
events and the frequency of negative events. We
conclude with a discussion of implications and applications of nonlinear
methods in political science research.
}

@@

The cohesiveness and strength of social movements in the pre-Internet era could be
operationalized by the number of people turning out to protest and through the
political events that precipitated and resulted from social gatherings \emph{en masse}.
The content of placards and protest
signs provided the protest narrative in their own words. Cohesive
social movements were reflected in their cohesive placards and posters. For
example, the \enquote{I am a man} posters held by the men involved in the 1968
sanitation workers' strike reflected their internal cohesion and unified,
goal-oriented movement (Windsor, 2020).

@@

Disentangling the complicated relationship between mass online mobilization---and
social mobilization in person---continues to be relevant, even as our modes
of communication change. The
political processes that transpired during and after the Arab Spring are still
in the process of being understood. As Zeitzoff (2017) recognizes,
social media is changing conflict, but the mechanisms and relationships by
which it does so are still being understood.

@@

Others have similarly expressed hesitation about the role of online
organizing. Some have argued that real-life strong ties outweigh online weak ties and
that the power of the strong ties through in-person contact is critical
to recruiting people for higher-risk activities
(Gladwell, 2010; McAdam, 1986).

@@

Citizens communicate with each other with increased transparency due to the
ubiquitous availability of social media such as low-cost and egalitarian
platforms like Twitter. Citizens are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and military
incursions, the degree of state repression, and the number of casualties. In
turn, the government---interested in remaining in power---monitors social media
platforms to ascertain emerging threats (Comninos, 2011),
such as where new groups are forming
and where existing groups are strengthening, where gatherings are held, and who
the idea entrepreneurs are. An example of a leader
expressing his interest in
such public conversation is given by Bashar al-Assad in a speech on June 3, 2012:

@@

With few exceptions, Assad chose to keep social media platforms up and running.
Gohdes (2015) notes that regimes can attempt to thwart the growth of social
protest online and in the streets by shutting down the Internet.
During these times, however, she finds that military repression increases.
However, government information about the citizens decreases; as such, many
regimes have elected to maintain Internet connectivity as a means of monitoring
and surveilling their citizens.

@@

Previous research on social movements supports the idea that participants are
likely to join movements when they know and have a personal connection to
someone else already involved (McAdam, 1986). The type of collective
contentious behavior that manifests in society is a function of the available
political opportunity structures. The political opportunity structure are
\enquote{comprised \emph{(sic)} of specific configuration of resources, institutional
arrangements and historical precedents for social mobilization, which facilitate
the development of protest movements in some instances and constrain them in
others} ({\textbf{???}}). What emerges as strategies and tactics in contentious
politics are, in essence, a function of what the system makes possible. This
includes the number of people and their disposition (i.e.~nonviolent or
violent), as well as their internal structure. The more cohesive a group is, the
greater threat they pose to the establishment.

@@

Steinert-Threlkeld (2017) modeled a similar relationship
between social media and social mobilization alongside event data. This
groundbreaking work demonstrates that participants on the fringe of a social
movement contribute substantially to the overall level of protest. In addition
to providing new methods to this growing field of study, we build on this
previous work by demonstrating how sociolinguistic cohesion evolves in
response to negative and positive exogenous events.

@@

The corpus of tweets used included 697428 Syrian tweets
occurring between March 30--June 15, 2012. Using the Twitter streaming API,
servers at the University of Texas at Austin collected the tweets. The servers
were run on a stable (LTS) version of Ubuntu Linus (v. 10.04) and were connected
directly to the high-bandwidth UT network maintained by IT services.

@@

The social cohesion calculation was performed on the tweet corpus containing
both English and Arabic tweets. All hashtags and URLs were included in the
analysis.

@@

@@

For WCRQA analyses, we decided to focus on RR and DET.

@@

>>>>

\abstract{
In recent years, individuals involved in social movements have taken to social
media platforms to reach broader audiences at rapid speeds. Although
micro-blogging has become a primary means of information dissemination during
times of turmoil, the degree to which virtual mobilization reflects or drives
real-world movements is unclear. Here, we explore the dynamics of real-world
events and Twitter social cohesion in Syria during the Arab Spring. Using the
nonlinear methods cross-recurrence quantification analysis and windowed
cross-recurrence quantification analysis, we investigate if frequency of
events of different intensities are coupled with social cohesion found in
Syrian tweets. Results indicate that virtual social cohesion is coupled with
the counts of all and negative events each day but shows a more nuanced
connection to positive events, suggesting that virtual mobilization does
reflect real-world movements. We conclude with a discussion of implications
and applications of nonlinear methods in political science research.
}

@@

The cohesiveness and strength of social movements in the pre-Internet era could
be operationalized by the number of people turning out to protest and through
the political events that precipitated and resulted from social gatherings \emph{en
masse}. Cohesive social movements were reflected in their cohesive placards and
posters. For example, the \enquote{I am a man} posters held by the men involved in the
1968 sanitation workers' strike reflected their internal cohesion and unified,
goal-oriented movement (Windsor, 2020).

@@

Disentangling the complicated relationship between mass online
mobilization---and social mobilization in person---continues to be relevant,
even as our modes of communication change. The political events that
transpired during and after the Arab Spring are still in the process of being
understood. As Zeitzoff (2017) recognizes, social media is
changing conflict, but the mechanisms and relationships by which it does so are
still being understood.

@@

\noindent Others have similarly expressed hesitation about the role of online
organizing. Some have argued that real-life strong ties outweigh online weak ties and
that the power of the strong ties through in-person contact is critical
to recruiting people for higher-risk activities
(Gladwell, 2010; McAdam, 1986).

@@

Citizens communicate with each other with increased transparency due to the
ubiquitous availability of social media such as low-cost and egalitarian
platforms like Twitter. Citizens are able to access these tools via mobile
phones and computers with Internet connectivity, providing a stream of real-time
shared information. Citizens can share the location of events like protests and
military incursions, the degree of state repression, and the number of
casualties. In turn, the government---interested in remaining in
power---monitors social media platforms to ascertain emerging threats
(Comninos, 2011), such as where new groups are forming and where existing
groups are strengthening, where gatherings are held, and who the idea
entrepreneurs are.

An example of a leader expressing his interest in such public
conversation is given by Bashar al-Assad in a speech on June 3, 2012:

@@

\noindent With few exceptions, Assad chose to keep social media platforms up and running.
Gohdes (2015) notes that regimes can attempt to thwart the growth
of social protest online and in the streets by shutting down the Internet.
During these times, however, she finds that military repression increases.
However, government information about the citizens decreases; as such, many
regimes have elected to maintain Internet connectivity as a means of monitoring
and surveilling their citizens.

@@

Previous research on social movements supports the idea that participants are
likely to join movements when they know and have a personal connection to
someone else already involved (McAdam, 1986). The type of collective
contentious behavior that manifests in society is a function of the available
political opportunity structures. The political opportunity structure are
\enquote{comprised of \emph{(sic)} specific configuration of resources, institutional
arrangements and historical precedents for social mobilization, which facilitate
the development of protest movements in some instances and constrain them in
others} (Kitschelt, 1986). What emerges as strategies and tactics in
contentious politics are, in essence, a function of what the system makes
possible. This includes the number of people and their disposition (i.e.
nonviolent or violent), as well as their internal structure. The more cohesive a
group is, the greater threat they pose to the establishment.

@@

Steinert-Threlkeld (2017) modeled a similar relationship
between social media and social mobilization alongside event data. That
groundbreaking work demonstrated that participants on the fringe of a social
movement contribute substantially to the overall level of protest. In addition
to providing new methods to this growing field of study, we build on this
previous work by demonstrating how sociolinguistic cohesion evolves in
response to negative and positive exogenous events.

@@

The corpus of tweets used included 0 Syrian tweets
occurring between March 30--June 15, 2012. Using the Twitter streaming API,
servers at the University of Texas at Austin collected the tweets. The servers
were run on a stable (LTS) version of Ubuntu Linus (v. 10.04) and were connected
directly to the high-bandwidth UT network maintained by IT services.

@@

\noindent The social cohesion calculation was performed on the tweet corpus
containing both English and Arabic tweets. All hashtags and URLs were included
in the analysis. The raw social cohesion time series can be seen in Figure
\ref{fig:raw-cohesion-ts}.

\begin{figure}
\includegraphics[width=6.5in]{./results/primary/raw-cohesion-ts} \caption{Raw social cohesion metric time series.}\label{fig:raw-cohesion-ts}
\end{figure}

@@

The raw event time series for target filtered data can be seen in Figure
\ref{fig:raw-event-ts-target}. The raw event time series for source and target
filtered data can be seen in Figure \ref{fig:raw-event-ts-source-target}.

\begin{figure}
\includegraphics[width=6.5in]{./results/primary/raw-event-ts-target} \caption{Raw event count time series for target filtered data.}\label{fig:raw-event-ts-target}
\end{figure}

\begin{figure}
\includegraphics[width=6.5in]{./results/primary/raw-event-ts-source-target} \caption{Raw event count time series for source and target filtered data.}\label{fig:raw-event-ts-source-target}
\end{figure}

@@

For WCRQA analyses, we decided to focus on RR and DET, but similar analyses and
visualizations can be performed for any RQA metric.

@@

\leavevmode\hypertarget{ref-kitschelt1986political}{}%
Kitschelt, H. P. (1986). Political opportunity structures and political protest: Anti-nuclear movements in four democracies. \emph{British Journal of Political Science}, \emph{16}(1), 57--85.
